Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-05 resubmission/02-antecedents.qmd
review antecedentes jc (falta merit)

We adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions. This distinction is essential, as it captures different pathways through which individuals justify inequality. Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility. A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed. Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023]. The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions. For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023]. This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage. Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.

@@

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024]. This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time. While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research. Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.

>>>>

We adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions. This distinction is essential, as it captures different pathways through which individuals justify inequality. Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility. A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed. Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023]. The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions. For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023].

@@

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success would be more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024]. This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research. Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.
